---
title: "Data Analysis using R"
output: html_notebook
---

### Import necessary packages

I will first import all the necessary packages. As I'll do data analysis, I'll use R built in functions and also `ggplot` to plot certain graphs about the data to get better insights into it.

```{r}
install.packages('ggplot2')
library('ggplot2')
```

### Import data

The next step is to import the data that we will work with. I've taken the dataset from [Kaggle](https://www.kaggle.com/joniarroba/noshowappointments/download) about **medical apppointments** with 13 features and a lable describing whether the person showed up for the appointment or not.

```{r}
medical.dataset <- read.csv('data/dataset.csv', header = TRUE, sep = ',')
head(medical.dataset)
```

The dataset is now loaded into the notebook. Let's start exploring more information about the data.

### Exploratory data analysis

Let's start the exploratory analysis of the dataset.

#### Dataset exploration

Let's explore the information about the datset.

```{r}
print(paste("Rows: ", nrow(medical.dataset)))
print(paste("Columns: ", ncol(medical.dataset)))
```

The dataset has 110,527 rows and 14 columns out of which we know we have 13 features and 1 target.

```{r}
names(medical.dataset)
```

We have 13 features, let's explore them furthermore:

1. "PatientId": It is a unique identifier for each patient but would not be useful for any predictions.
2. "AppointmentId: It's a unique identifier for each appointment.
3. "Gender": Whether the person is a Female denoted by F or Male denoted by M.
4. "ScheduleDay": The day on which appointment was scheduled.
5. "AppointmentDay": The day of the appointment.
6. "Age": Age of the person.
7. "Neighbourhood": The neighbourhood to which the person belongs.
8. "Scholarship": Whether the person had scholarship (defined [Wikipedia](https://en.wikipedia.org/wiki/Bolsa_Fam%C3%ADlia)).
9. "Hipertension": Whether the person has hyptertension.
10. "Diabeletes": Whether the person has diabetes.
11. "Alcoholism": Whether the person is alcoholic.
12. "Handcap": Whether the person is physically challenged.
13. "SMS_received": Whether the person received a text message about the appointment.

The target variable is:

1. "No.show": Depicts whether the person didn't show up for the appointment.

```{r}
summary(medical.dataset)
```

The `summary` fuinction gives a really descriptive summary of what the dataset includes but I find using plots more useful and more understandable to get the essence of the dataset.
